Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-04 Paper/manuscript/manuscript_05.tex
edited intro

Gaussian processes (GPs) are natural (Paul: what does "natural" imply and as compared to what?) and flexible non-parametric models to define probability distributions over multi-dimensional functions \citep{rasmussen2006gaussian,neal1997monte}. This allows applying probabilistic inference to estimate functional relationships between one or more covariates and a single response variable. The defining feature of GPs is that the function values are assumed to jointly follow a multivariate normal distribution, which is characterized primarily by a covariance function, sometimes also referred to as covariance kernel. The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Several types of covariance functions with varying properties can be used, and they may even be combined for further increased flexibility. GPs are often referred to as non-parametric models because the number of parameters is not fixed, but rather increases with the number of data points, which allows to adapt model complexity to the data. As such, the term 'non-parametric' does not imply having no parameters but actually having a lot of them. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

One of the main limitations of exact GPs is that computational demands and memory requirements scale as $O(n^3)$ and $O(n^2)$, respectively, with the number $n$ of observations in the data used to fit the model. This effectively limits their application to rather small data sets of a few thousand observations at most. This problem becomes especially severe when performing full Bayesian inference via sampling methods, where in each sampling step we need to invert the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed, which we will briefly review in the following.

Sparse GPs are based on building low-rank approximations of the covariance matrix of the complete data by means of selecting a subset $m$ of training points on which to base computation. They reduce the dimension of the posterior distribution thereby reducing the memory requirements to $O(nm)$ and computational complexity to $O(nm^2)$ for some $m < n$. This approach still requires Cholesky factorization although now based on a smaller covariance matrix. While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation). As such, reducing the amount of data while maintaining the correct posterior distribution is not a simple procedure. A unifying view on sparse GPs based on approximate generative methods is provided in \cite{quinonero2005unifying}, while a general review can be found in \cite{rasmussen2006gaussian}. New developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

(Paul: This whole paragraph is unclear to me. Are all of these approaches approximations via splines, or are the splines just special cases of a more general method? In any case, I think the paragraph needs substantial revision). Other global approach make use of the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}. It is well-known (see, e.g., \cite{wahba1990spline}) that spline smoothing can be seen as Gaussian process regression with a specific choice of covariance function, but they do not generally have correspondence with GP models, and they do not have the flexibility of choosing different covariance functions as model structure. The Sparse Spectrum GP is based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving,gal2015improving}. Recently \citep{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat´ern class of kernels. Random Fourier Features \citep{rahimi2008random,rahimi2009weighted} is a method for approximating kernels.

Yet another approach to approximate GPs -- on which we will focus in the present paper -- was proposed by \cite{solin2018hilbert} who suggested to use reduced-rank approximations of the covariance kernel. This method decomposes the kernel into basis functions which represent the GP as a linear model and thus makes its estimation considerably faster. In principal, we are free in the choice of basis functions, but the Laplace eigenfunctions were found to be particularly appealing \citep{solin2018hilbert}. Not only can they be computed analytically, but they are also independent of the particular choice of the covariance kernel including the hyperparameters. The latter comes with several advantages when performing inference via algorithms that require differentiation of the posterior density (Paul: cite something here). In a fully Bayesian framework based on sampling methods, the proposed approximate GP model has a computational complexity of just $O(nm+m)$, with $m<<n$ in every sampling step. The posterior distribution of the reduced-rank GP approximation is of dimension $m$ and thus considerably smaller than in the case of exact GPs. This not only lowers the memory requirements and computational complexity but also reduces the correlations between latent values and the corresponding hyperparameters of the GP, which can help to further improve sampling efficiency.

While \cite{solin2018hilbert} fully developed the mathematical theory behind reduced-rank approximations of GPs, they did not put much effort in analyzing the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. However, for practical implementations of approximate GPs, the latter factors are critical and thus require further investigation. In the present paper, we will fill this gap by providing practical recommendations for the choice of these factors based on the recognized relations among them. What is more, we provide intuitive visualizations of these relations that will help users to improve performance and save computation time, while at the same time maintaining close approximations of exact GPs.

Although there are several GP specific software packages available to date (Paul: cite those. Gabi: This full paragraph comes from Aki. I think he refers to some Splines methods and maybe others approaches as well), each provide efficient implementations only for a restricted range of GP based models. In this paper we do not focus on the fastest possible inference for a specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks such as Stan \citep{carpenter2017stan}. To this end, we want to make full Bayesian inference over Gaussian process models accessible to a broader audience by making them easy, fast and save to apply.

>>>>

Gaussian processes (GPs) are flexible define probability distributions over multi-dimensional non-linear functions \citep{rasmussen2006gaussian,neal1997monte}. GPs are defind by a mean and covariance function and any finite set of function values are jointly distributed as a multivariate normal.  The covariance function encodes our prior assumptions about the functional relationship, such as continuity, smoothness, periodicity and scale properties. GPs not only allow for non-linear effects but can also implicitly handle interactions between covariates. Different types of covariance functions can be used, and they may be combined for further increased flexibility. Due to their generality and flexibility, GPs are of broad interest across machine learning and statistics \citep{rasmussen2006gaussian,neal1997monte}. Among others, they find application in the fields of spatial epidemiology \citep{diggle2013statistical,carlin2014hierarchical}, robotics and control \citep{deisenroth2015gaussian}, signal processing \citep{sarkka2013spatiotemporal}, as well as Bayesian optimization and probabilistic numerics \citep{roberts2010bayesian,briol2015probabilistic,hennig2015probabilistic}.

Given $n$ observations in the data, the computational cost and memory requirements of exact GP implementation in general scales as $O(n^3)$ and $O(n^2)$, respectively. This would limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes especially severe when performing full Bayesian inference via sampling methods, where in each sampling step we have $O(n^3)$ computational cost when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed, which we will briefly review in the following.

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} can be presented with reduced memory cost $O(nm)$ and the corresponding computational complexity is $O(nm^2)$.
% While computational complexity is reduced considerably, ensuring convergence to the exact GP model is not necessarily straightforward (Paul: this may need a citation).
A unifying view on sparse GPs based on approximate generative methods
is provided in \cite{quinonero2005unifying}, while a general review
can be found in \cite{rasmussen2006gaussian}.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

% (Paul: This whole paragraph is unclear to me. Are all of these approaches approximations via splines, or are the splines just special cases of a more general method? In any case, I think the paragraph needs substantial revision).
Alternative low-rank approximation is based on forming a basis function approximation with $m \ll$. The basis functions are presented explicitly, but could also be used to form a low rank covariance matrix approximation. Common basis function  approximations are based on the spectral analysis and series expansions of Gaussian processes \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Certain spline smoothing basis functions are equivalent to GPs with certain covariance function \citep{wahba1990spline}, but not all spline models are GPs (sometimes splines are not marginalization consistent)-
%
The sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving,gal2015improving}. Recently \citep{hensman2017variational} presented a variational Fourier feature approximation for Gaussian processes that was derived for the Mat´ern class of kernels. Random fourier features \citep{rahimi2008random,rahimi2009weighted} is a method for approximating kernels.

In this paper we focus on Laplace eigenfunction basis function approximation for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches are computationally as linear models, which is attractive property in modular probabilistic programming models where there is a big benefit id approximation specific computation is simple. The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$ the computation cost when only the covariance function parameters change is $O(mn+m)$ (or $O(m^3)$ if marginalization can be used). This is a big advantage for iterative algorithms such as Markov chain Monte Carlo (MCMC). The further advantage is the reduced memory cost in automatic differentation used in modern probabilistic programming frameworks, as the automatic differentation memory cost scales as the computational cost and not as the usual memory cost for the posterior density computation. The basis function approach provides also easy way to use non-centered parameterization reducing posterior dependency between parameters presenting the function and the covariance function parameters, which helps to further improve MCMC efficiency.

While \citet{solin2018hilbert} fully developed the mathematical theory behind the specific reduced-rank approximation, for practical implementations of approximate GPs in probabilistic programming framework further work is needed. In this paper we analyze in detail the performance and accuracy of the method in relation to key factors such as the number of basis functions, desired prediction space, or properties of the true functional relationship between covariates and response variable. We provide intuitive visualizations and practical recommendations for the choice of these factors, which will help users to improve computational performance while maintaining close approximation to exact GPs.

Although there are several GP specific software packages available to date (GPML, GPstuff, GPy, GPflow), each provide efficient implementations only for a restricted range of GP based models. In this paper we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks such as Stan \citep{carpenter2017stan}.
